---
title: "Bayesian Inference with MCMC"
subtitle: "odin2/dust2/monty Vignette 06 (R)"
format:
  html:
    toc: true
    code-fold: false
  gfm:
    toc: true
  pdf:
    toc: true
---

## Introduction

Full Bayesian inference with particle filter + MCMC using odin2/dust2/monty.

The Julia companion (parent directory) uses the same model, data, and priors
but with ModelingToolkit.jl + Turing.jl (deterministic ODE + NUTS).

## Model Definition

```{r}
library(odin2)
library(dust2)
library(monty)

sir <- odin({
  update(S) <- S - n_SI
  update(I) <- I + n_SI - n_IR
  update(R) <- R + n_IR
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(incidence, zero_every = 1) <- 0
  update(incidence) <- incidence + n_SI

  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IR <- 1 - exp(-gamma * dt)
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)

  cases <- data()
  cases ~ Poisson(incidence + 1e-6)

  beta <- parameter(0.5)
  gamma <- parameter(0.1)
  I0 <- parameter(10)
  N <- parameter(1000)
})
```

## Generate Synthetic Data

```{r}
true_pars <- list(beta = 0.5, gamma = 0.1, I0 = 10, N = 1000)
times <- seq(0, 50, by = 1)

sys <- dust_system_create(sir, true_pars, dt = 1, seed = 1)
dust_system_set_state_initial(sys)
obs_result <- dust_system_simulate(sys, times)
observed <- round(obs_result[4, -1])

data <- data.frame(time = times[-1], cases = observed)
cat("Observed cases (first 10):", head(observed, 10), "\n")
cat("Total observed:", sum(observed), "\n")
```

## Set Up Inference

Priors: β ~ Gamma(shape=2, rate=4), γ ~ Gamma(shape=2, rate=20). These are identical to the Julia Turing priors.

```{r}
filter <- dust_filter_create(sir, time_start = 0, data = data,
                             n_particles = 200, seed = 42)

packer <- monty_packer(c("beta", "gamma"),
                       fixed = list(I0 = 10, N = 1000))

likelihood <- dust_likelihood_monty(filter, packer)

prior <- monty_dsl({
  beta ~ Gamma(shape = 2, rate = 4)
  gamma ~ Gamma(shape = 2, rate = 20)
})

posterior <- likelihood + prior
```

## Run MCMC

We use the random walk Metropolis sampler (2000 iterations).

```{r}
vcv <- matrix(c(0.005, 0, 0, 0.001), 2, 2)
sampler <- monty_sampler_random_walk(vcv)

timing <- system.time({
  samples <- monty_sample(posterior, sampler, 2000, initial = c(0.4, 0.08))
})
elapsed <- timing["elapsed"]
cat("Sampling complete: 2000 iterations in", round(elapsed, 2), "s\n")
```

## Results

```{r}
burnin <- 500
beta_samples <- samples$pars[1, (burnin + 1):2000, 1]
gamma_samples <- samples$pars[2, (burnin + 1):2000, 1]

cat("β: mean =", round(mean(beta_samples), 3),
    ", 95% CI = [", round(quantile(beta_samples, 0.025), 3),
    ",", round(quantile(beta_samples, 0.975), 3), "]\n")
cat("γ: mean =", round(mean(gamma_samples), 3),
    ", 95% CI = [", round(quantile(gamma_samples, 0.025), 3),
    ",", round(quantile(gamma_samples, 0.975), 3), "]\n")
cat("True: β = 0.5, γ = 0.1\n")
```

## Posterior Histograms

```{r fig.width=7, fig.height=3.5}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

hist(beta_samples, breaks = 30, main = "Posterior: β", xlab = "β",
     probability = TRUE, col = rgb(0.27, 0.51, 0.71, 0.7), border = NA)
abline(v = 0.5, col = "red", lwd = 2)
legend("topright", legend = "True", col = "red", lwd = 2, bty = "n")

hist(gamma_samples, breaks = 30, main = "Posterior: γ", xlab = "γ",
     probability = TRUE, col = rgb(0.27, 0.51, 0.71, 0.7), border = NA)
abline(v = 0.1, col = "red", lwd = 2)
legend("topright", legend = "True", col = "red", lwd = 2, bty = "n")
```

## Trace Plots

```{r fig.width=7, fig.height=4}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

plot(seq_along(beta_samples), beta_samples, type = "l",
     col = rgb(0.27, 0.51, 0.71), lwd = 0.5,
     xlab = "Iteration", ylab = "β", main = "Trace: β")
abline(h = 0.5, col = "red", lwd = 2)

plot(seq_along(gamma_samples), gamma_samples, type = "l",
     col = rgb(0.27, 0.51, 0.71), lwd = 0.5,
     xlab = "Iteration", ylab = "γ", main = "Trace: γ")
abline(h = 0.1, col = "red", lwd = 2)
```

## Effective Sample Size

```{r}
ess_simple <- function(x) {
  n <- length(x)
  max_lag <- min(n - 1, 1000)
  mu <- mean(x)
  var0 <- sum((x - mu)^2) / n
  if (var0 == 0) return(n)
  rho <- numeric(max_lag)
  for (k in seq_len(max_lag)) {
    rho[k] <- sum((x[1:(n - k)] - mu) * (x[(k + 1):n] - mu)) / (n * var0)
  }
  tau <- 1
  for (k in seq(1, max_lag - 1, by = 2)) {
    pair <- rho[k] + rho[k + 1]
    if (pair < 0) break
    tau <- tau + 2 * pair
  }
  return(n / tau)
}

ess_beta <- ess_simple(beta_samples)
ess_gamma <- ess_simple(gamma_samples)

cat("── Performance ──\n")
cat("Sampler:    Random walk (monty)\n")
cat("Iterations: 2000 (burn-in:", burnin, ")\n")
cat("Time:      ", round(elapsed, 2), "s\n")
cat("ESS (β):   ", round(ess_beta, 1), "\n")
cat("ESS (γ):   ", round(ess_gamma, 1), "\n")
cat("ESS/s (β): ", round(ess_beta / elapsed, 1), "\n")
cat("ESS/s (γ): ", round(ess_gamma / elapsed, 1), "\n")
```
