---
title: "Bayesian Inference with odin2/dust2/monty"
subtitle: "R companion for vignette 07"
format:
  html:
    code-fold: false
  gfm:
    toc: true
  pdf:
    toc: true
---

## Overview

This R companion demonstrates Bayesian inference for the SIR model using
odin2/dust2/monty. Since R's odin2 does not currently expose adjoint gradients
through its public API, only random-walk MCMC is available (not HMC/NUTS).

```{r}
library(odin2)
library(dust2)
library(monty)
```

## Model and Data

```{r}
sir <- odin({
  deriv(S) <- -beta * S * I / N
  deriv(I) <- beta * S * I / N - gamma * I
  deriv(R) <- gamma * I
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  beta <- parameter(0.5)
  gamma <- parameter(0.1)
  I0 <- parameter(10)
  N <- parameter(1000)
  obs <- data()
  obs ~ Poisson(max(I, 1e-6))
})
```

```{r}
pars <- list(beta = 0.5, gamma = 0.1, I0 = 10, N = 1000)
sys <- dust_system_create(sir, pars,
  ode_control = dust_ode_control(atol = 1e-8, rtol = 1e-8))
dust_system_set_state_initial(sys)
times <- 1:50
state <- dust_system_simulate(sys, times)

set.seed(1)
I_vals <- state[2, ]
obs <- rpois(length(times), pmax(I_vals, 1e-6))
data <- data.frame(time = times, obs = obs)

cat("Generated", length(obs), "observations, range:",
    range(obs), "\n")
```

## ESS Helper

```{r}
ess_ipse <- function(x) {
  n <- length(x)
  mu <- mean(x)
  var0 <- sum((x - mu)^2) / n
  if (var0 == 0) return(n)
  max_lag <- min(n - 1, 1000)
  rho <- sapply(1:max_lag, function(k) {
    sum((x[1:(n - k)] - mu) * (x[(k + 1):n] - mu)) / (n * var0)
  })
  tau <- 1
  for (k in seq(1, max_lag - 1, by = 2)) {
    ps <- rho[k] + rho[k + 1]
    if (ps < 0) break
    tau <- tau + 2 * ps
  }
  n / tau
}
```

## Inference: Random-Walk MCMC

R odin2 does not currently provide gradients via its public API,
so HMC/NUTS sampling is not available. We use random-walk Metropolis–Hastings.

```{r}
d <- dust_unfilter_create(sir, data = data, time_start = 0,
  ode_control = dust_ode_control(atol = 1e-8, rtol = 1e-8))
packer <- monty_packer(c("beta", "gamma"),
  fixed = list(I0 = 10, N = 1000))
ll <- dust_likelihood_monty(d, packer)

prior <- monty_dsl({
  beta ~ Gamma(shape = 2, rate = 4)
  gamma ~ Gamma(shape = 2, rate = 20)
})
posterior <- ll + prior
```

```{r}
vcv <- diag(c(0.0001, 0.000004))
sampler <- monty_sampler_random_walk(vcv)
```

```{r}
# Warmup
invisible(monty_sample(posterior, sampler, 200,
  initial = matrix(c(0.5, 0.1), ncol = 1)))

# Timed run
t0 <- proc.time()
samples <- monty_sample(posterior, sampler, 5000,
  initial = matrix(c(0.5, 0.1), ncol = 1))
elapsed <- (proc.time() - t0)[3]

burnin <- 1000
bs <- samples$pars[1, (burnin + 1):5000, 1]
gs <- samples$pars[2, (burnin + 1):5000, 1]
```

## Results

```{r}
cat(sprintf("R odin2 Random-Walk MCMC\n"))
cat(sprintf("  Time:     %.2fs\n", elapsed))
cat(sprintf("  β:        %.4f ± %.4f\n", mean(bs), sd(bs)))
cat(sprintf("  γ:        %.4f ± %.4f\n", mean(gs), sd(gs)))
cat(sprintf("  ESS(β):   %.1f\n", ess_ipse(bs)))
cat(sprintf("  ESS(γ):   %.1f\n", ess_ipse(gs)))
cat(sprintf("  ESS/s(β): %.1f\n", ess_ipse(bs) / elapsed))
cat(sprintf("  ESS/s(γ): %.1f\n", ess_ipse(gs) / elapsed))
cat(sprintf("  iter/s:   %.0f\n", 5000 / elapsed))
```

```{r}
#| label: fig-posteriors
#| fig-cap: "Posterior distributions (R odin2 RW-MCMC)"
#| fig-width: 8
#| fig-height: 4
par(mfrow = c(1, 2))

hist(bs, breaks = 40, col = "steelblue", border = "white",
     main = expression("Posterior: " * beta),
     xlab = expression(beta), probability = TRUE)
abline(v = 0.5, col = "red", lty = 2, lwd = 2)

hist(gs, breaks = 40, col = "steelblue", border = "white",
     main = expression("Posterior: " * gamma),
     xlab = expression(gamma), probability = TRUE)
abline(v = 0.1, col = "red", lty = 2, lwd = 2)
```

## Notes

The R odin2/dust2/monty stack uses compiled C++ for the ODE solver and likelihood
evaluation, making random-walk MCMC very fast (~12,500 iter/s). However, without
gradient information, HMC/NUTS cannot be used.

The Julia MTK approach in vignette 07 demonstrates how symbolic sensitivity equations
can provide exact gradients, enabling NUTS sampling with ~1,200 iter/s and much
higher effective sample sizes per iteration.
